---
title: "Metastudy: activation of the mTOR signalling pathway in bone marrow-derived macrophages and microglia"
output: html_notebook
---

```{r setup}
require("knitr")
opts_knit$set(root.dir = normalizePath('..'))

library(edgeR)
library(DESeq2)
library(VennDiagram)
library(vsn)
library("dplyr")
library("ggplot2")
library("reshape2")

```

# Outline

* When the blood brain barrier is disrupted, bone marrow-derived macrophages (BMDM) can migrate across into the brain tissue.
* Disruption can be caused by a tumour or irradiation.
* They join the existing population of microglia (MG).
* The two cell types are difficult to distinguish.
* There is epigenetic evidence that certain pathways do differ between BMDM and MG.
* Both cell types seem to acquire a new 'education' in a tumour environment. They are known as tumour-associated (TA).

# Hypothesis
*The mTOR pathway is more active in TA-MG (relative to healthy MG) than in TA-BMDM (relative to healthy BMDM).*

# Available data

Two public mouse RNA-Seq datasets retrieved from GEO.

GSE68376: _Genetic Cell Ablation Reveals Clusters of Local Self-Renewing Microglia in the Mammalian Central Nervous System_ Bruttger, J. et al. Immunity (2015).  All samples from a healthy animal.  3 x BMDM biological replicates; 2 x WT MG biological replicates; 3 x repopulating MG biological replicates (following genetic ablation).

GSE86573: _Macrophage Ontogeny Underlies Differences in Tumor-Specific Education in Brain Malignancies_ Bowman et al. Cell Reports (2016). 3 x BMDM biological replicates and 3 x MG biological replicates from 2 different mouse models (total of 12 samples, 6 of each cell type). 5 normal blood monocytes, 3 normal MG (8 control samples).

I'll only use *GSE86573* for this analysis. Healthy blood monocytes are treated as the control for TA-BMDM.

# Aim

Use differential expression analysis of the 'gene count' data to test the hypothesis. Identify significantly DE genes in TA-BMDM vs healthy monocytes and TA-MG vs healthy MG. Then find genes that only appear in one or the other list. Finally, look for enrichment in mTOR pathway.

# Analysis

Import necessary libraries and load data. We set the metadata here too. We'll set the FDR and minimum effect size (absolute fold change) here for all comparisons. I've added a 'composite' column to the metadata, which will make it easier to define comparisons in the analysis.

```{r, message=FALSE, warning=FALSE, results="hide"}
fdr <- 0.01  # Conflicting advice on this value in the paper: either 0.01 or 0.05
logFC <- 1  # FC = 2

source('io/output.R')
source('_settings.R')

in.file <- file.path(data.dir.raid, "rnaseq", "GSE86573", "GSE86573_raw_counts.csv")
data.bowman <- read.csv(in.file, row.names = 1)
# we'll change the sample names for clarity
colnames(data.bowman) <- c("gl261_bmdm_rep1", "gl261_mg_rep1", "gl261_bmdm_rep2", "gl261_mg_rep2", "gl261_bmdm_rep3", "gl261_mg_rep3",
                    paste0("monocyte_rep", seq(1:5)), paste0("mg_rep", seq(1:3)),
                    "gemm_bmdm_rep1", "gemm_mg_rep1", "gemm_bmdm_rep2", "gemm_mg_rep2", "gemm_bmdm_rep3", "gemm_mg_rep3")

# we don't use the control MG samples here
# data.bowman <- data.bowman[, !grepl("^mg_rep", colnames(data.bowman))]

meta.bowman <- data.frame(row.names = colnames(data.bowman))

meta.bowman$cell_type <- "monocyte"
meta.bowman$cell_type[grepl("mg", rownames(meta.bowman))] <- "mg"

meta.bowman$model <- "healthy"
meta.bowman$model[grepl("gemm", rownames(meta.bowman))] <- "gemm"
meta.bowman$model[grepl("gl261", rownames(meta.bowman))] <- "gl261"

meta.bowman$disease <- "healthy"
meta.bowman$disease[grepl("gemm", rownames(meta.bowman))] <- "tumour"
meta.bowman$disease[grepl("gl261", rownames(meta.bowman))] <- "tumour"

meta.bowman$group <- paste(meta.bowman[,1], meta.bowman[,2], meta.bowman[, 3], sep='_')
# healthy_healthy is a mess so replace it
meta.bowman$group <- as.factor(sub('healthy_healthy', 'healthy', meta.bowman$group))

data <- data.bowman
meta <- meta.bowman
meta
```

```{r, eval=FALSE, echo=FALSE}
#' Here is how we would load the other dataset, if we required it.

in.file <- file.path(data.dir.raid, "rnaseq", "GSE68376", "GSE68376_normalizedCounts.txt")
data.bruttger <- read.csv(in.file, row.names = 1, sep='\t')
colnames(data.bruttger) <- c(
  paste0("ctrl_bmdm_rep", seq(1:3)), paste0("ctrl_mg_rep", seq(1:2)), paste0("repop_mg_rep", seq(1:3))
)

# Think that the authors only used the repopulated MG cells?
# data.bruttger <- data.bruttger[, !grepl('ctrl_mg', colnames(data.bruttger))]

meta.bruttger <- data.frame(row.names = colnames(data.bruttger))

meta.bruttger$cell_type <- "bmdm"
meta.bruttger$cell_type[grepl("mg", rownames(meta.bruttger))] <- "mg"

meta.bruttger$model <- "endog"
meta.bruttger$disease <- "healthy"

meta.bruttger$group <- paste(meta.bruttger[,1], meta.bruttger[, 3], sep='_')

# Combine all data together.

genes <- intersect(rownames(data.bowman), rownames(data.bruttger))

meta <- rbind(meta.bowman, meta.bruttger)
data <- cbind(data.bowman[genes,], data.bruttger[genes,])

meta$group <- as.factor(paste(as.vector(meta[,1]), as.vector(meta[, 3]), sep='_'))
meta$composite <- paste(as.vector(meta[,1]), as.vector(meta[, 2]), as.vector(meta[, 3]), sep='_')
```

Normalise the data and look at the number of genes. 

```{r}
dgel <- DGEList(counts = data, samples = meta)
dgel <- calcNormFactors(dgel)
sprintf("Number of genes in raw data: %i", nrow(dgel))
```

Some of these have very low expression. Filtering them out now will help reduce the number of hypothesis tests we will carry out later, and increase the power accordingly. We should use counts per million (CPM) so that we take library size into account. In the smallest sample (~5mi reads), 1 CPM means 5 reads. In the largest sample (~30mi reads), 1 CPM means 30 reads. Let's require that at least 2 samples have CPM > 1 to retain a gene.

```{r}
keep <- rowSums(cpm(dgel) > 1) >= 2
dgel <- dgel[keep, ,  keep.lib.sizes=F]
sprintf("Number of genes in filtered data: %i. %.2f%% of genes retained.", nrow(dgel), sum(keep) / length(keep) * 100)
```

Let's look at the MDS and PCA plots. These can indicate batch effects and/or significant groupings.

```{r}
plotMDS(dgel)
title("MDS plot")

dds <- DESeqDataSetFromMatrix(countData = round(data), colData = meta, design = ~0 + group)
dds <- estimateSizeFactors(dds)
vsd <- vst(dds, blind = FALSE)

plotPCA(vsd, intgroup = c("cell_type", "disease", "model"))
```

All the groups cluster together. If we remove monocytes, do we get a more apparent separation of the remaining samples?

```{r}
samples = rownames(meta[meta$cell_type != 'monocyte',])
dgel.reduced <- DGEList(counts = data[,samples], samples = meta[samples,])
plotMDS(dgel.reduced)

dds <- DESeqDataSetFromMatrix(countData = round(data[, samples]), colData = meta[samples,], design = ~0 + group)
dds <- estimateSizeFactors(dds)
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("cell_type", "disease", "model"))
```

Yes - now the first dimension separates the two models and endogenous samples and the second dimension separates MG and BMDM.

Have toyed with removing monocytes, but leave them in for now. They may be used as the control for the BMDM population.

```{r, eval=F, echo=F}
data <- data[, samples]
meta <- droplevels(meta[samples,])  # drop unused factors
dgel <- DGEList(counts = data, samples = meta)
dgel <- calcNormFactors(dgel)
sprintf("EdgeR: Number of genes in raw data: %i", nrow(dgel))
keep <- rowSums(cpm(dgel) > 1) >= 2
dgel <- dgel[keep, ,  keep.lib.sizes=F]
sprintf("EdgeR: Number of genes in filtered data: %i. %.2f%% of genes retained.", nrow(dgel), sum(keep) / length(keep) * 100)
```

We can also transform our data for the purpose of running other exploratory methods like PCA or hierarchical clustering. We could just use $\log_2(x+1)$, but there are other options including `rlog` (slow, disabled for now) and `vst`. Here, we take two samples and look at how their gene values compare under the three transformations.

```{r}
dds <- DESeqDataSetFromMatrix(countData = round(data), colData = meta, design = ~0 + group)
dds <- estimateSizeFactors(dds)

# rld <- rlog(dds, blind = FALSE)
vsd <- vst(dds, blind = FALSE)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  # as_data_frame(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"))
  
colnames(df)[1:2] <- c("x", "y")  

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  

```


## `DESeq2`: recreating the results of Bowman et al.

In the published study by Bowman et al., the authors report the number of up- (Fig 2E) and downregulated (Fig S2D) genes in model-specific comparisons of BMDM vs MG. Can we approximately recreate those? The source code for their own analyses is available [here](https://bitbucket.org/bowmanr/joycelab-brain-tme). 

We don't filter lowly-expressed genes for this purpose, as it doesn't appear that the authors carried out this step in their analysis.


```{r}
samples <- colnames(data)
dds <- DESeqDataSetFromMatrix(countData = round(data[, samples]), colData = meta[samples,], design = ~group)
dds <- estimateSizeFactors(dds)
des <- DESeq(dds)

de.up = list()
de.down = list()

for (t in c("gemm", "gl261", "endog")) {
  if (t == 'endog') {
    u = "healthy"
  } else {
    u = paste0(c(t, "tumour"), collapse='_')
  }
  res <- results(des, contrast=list(paste0(c("groupmonocyte", u), collapse='_'), paste0(c("groupmg", u), collapse='_')), cooksCutoff = F)
  padj <- res$padj
  padj[is.na(padj)] = 1.
  print(sprintf(
    "Model %s. BMDM vs MG, FDR = %.02f, logFC cutoff %.1f. %i genes up. %i genes down.",
    t, fdr, logFC,
    sum((padj < fdr) & (res$log2FoldChange > logFC)),
    sum((padj < fdr) & (res$log2FoldChange < -logFC))
  ))
  de.up[[t]] <- res[(padj < fdr) & (res$log2FoldChange > logFC),]
  de.down[[t]] <- res[(padj < fdr) & (res$log2FoldChange < logFC),]
}

area1 = nrow(de.up[[1]])
area2 = nrow(de.up[[2]])
area3 = nrow(de.up[[3]])

g12 = intersect(
  rownames(de.up[[1]]), rownames(de.up[[2]])
)
n12 = length(g12)
n23 = length(intersect(
  rownames(de.up[[2]]), rownames(de.up[[3]])
))
n13 = length(intersect(
  rownames(de.up[[1]]), rownames(de.up[[3]])
))

g123 = intersect(
  intersect(
    rownames(de.up[[1]]), rownames(de.up[[2]])
  ),
  rownames(de.up[[3]])
)
n123 = length(g123)

grid.newpage()
venn <- draw.triple.venn(area1, area2, area3, n12, n23, n13, n123, category=names(de.up), fill = c("blue", "red", "white"))

# redefine g12: DE genes that overlap GEMM and GL261 but NOT healthy
g12 <- setdiff(g12, g123)

```

Adding the numbers in the Venn diagrams in the publication, here are the numbers reported:

* GEMM BMDM vs MG: 1059 up, 780 down.
* GL261 BMDM vs MG: 712 up, 660 down.
* Healthy BMDM vs MG: 840 up, 865 down.

We can ignore the final result, since the authors used the Bruttger data for this section. Otherwise, we are finding very similar results. The Venn diagram is quite close to Fig 2E. A quick check shows that many but not all of the genes listed are indeed detected. The discrepancy isn't surprising, because we've comapred against different control data.

```{r}


expctd_in_g12 = c("Ccl1", "Ccl17", "Ccl22", "Cd80", "Jak2", "Kmo", "Ptger2")
for (g in expctd_in_g12) {
  print(g %in% g12)
}

expctd_in_g123 = c("Ahr", "Cd40", "Ciita", "Pdgfc", "Pf4", "Pparg", "Vdr", "Vav3")
for (g in expctd_in_g123) {
  print(g %in% g123)
}
```


## `edgeR`: Which genes respond differently in TA-MG vs healthy MG? Which genes respond differently in TA-BMDM vs healthy monocytes?

Now we can analyse DE in `edgeR`. The most crucial part is defining the _design matrix_ and _contrasts_, which together encode the comparisons we want to make.

Here we only include the combined terms: TA-MG, TA-BMDM, H-MG, H-BMDM.

```{r}
dgel <- DGEList(counts = data, samples = meta)
design <- model.matrix(~0 + group, data = meta)
dgel <- estimateDisp(dgel, design)
fit <- glmFit(dgel, design)

lrt <- glmTreat(fit, contrast=c(1, 0, -1, 0, 0, 0), lfc=logFC)
gemm_mg.vs.h_mg <- topTags(lrt, n=Inf, p.value=fdr)  # p value here is the FDR

lrt <- glmTreat(fit, contrast=c(0, 1, -1, 0, 0, 0), lfc=logFC)
gl261_mg.vs.h_mg <- topTags(lrt, n=Inf, p.value=fdr)  # p value here is the FDR

lrt <- glmTreat(fit, contrast=c(0, 0, 0, 1, 0, -1), lfc=logFC)
gemm_mono.vs.h_mono <- topTags(lrt, n=Inf, p.value=fdr)  # p value here is the FDR

lrt <- glmTreat(fit, contrast=c(0, 0, 0, 0, 1, -1), lfc=logFC)
gl261_mono.vs.h_mono <- topTags(lrt, n=Inf, p.value=fdr)  # p value here is the FDR

announceDE <- function(dat, ttl) {
  sprintf(
    "** %s ** Found %i DE genes at FDR=%.03f. Of these, %i are down and %i are up",
    ttl,
    dim(dat)[1], 
    fdr,
    sum(dat$table$logFC < 0),
    sum(dat$table$logFC > 0)
  )  
}

announceDE(gemm_mg.vs.h_mg, "TA-MG vs healthy MG (GEMM model)")
announceDE(gl261_mg.vs.h_mg, "TA-MG vs healthy MG (GL261 model)")
announceDE(gemm_mono.vs.h_mono, "TA-BMDM vs healthy monocyte (GEMM model)")
announceDE(gl261_mono.vs.h_mono, "TA-BMDM vs healthy monocyte (GL261 model)")


```
Plot a Venn diagram showing all DE genes.
```{r}

plotVenn.2 <- function(dat1, dat2, category) {
  area.1 <- nrow(dat1$table)
  area.2 <- nrow(dat2$table)
  area.cross <- length(intersect(rownames(dat1), rownames(dat2)))
  grid.newpage()
  venn <- draw.pairwise.venn(
    area.1, area.2, area.cross, category = category,
    lty = rep("blank", 2), fill = c("blue", "red"), alpha = rep(0.5, 2)
  )
}

plotVenn.2(gemm_mg.vs.h_mg, gl261_mg.vs.h_mg, c("GEMM MG", "GL261 MG"))
plotVenn.2(gemm_mono.vs.h_mono, gl261_mono.vs.h_mono, c("GEMM BMDM", "GL261 BMDM"))


```

For my own records: this is _exactly_ the same as using interaction variables (`cell_type:disease`) and carrying out the fitting on that instead.

Look at the DE gene list for suggestion of enrichment in the mTOR pathway. We need to check each of the segments in the Venn diagram.

```{r}
mtor_genes <- c(
"Eif3h",
"Eif4ebp1",
"Hif1a",
"Pik3r5",
"Pld3",
"Prkca",
"Prr5l",
"Rhoc",
"Rps2",
"Rps5",
"Rps7",
"Rps8",
"Rps10",
"Rps12",
"Rps13",
"Rps15",
"Rps16",
"Rps17",
"Rps18",
"Rps19",
"Rps20",
"Rps21",
"Rps23",
"Rps24",
"Rps25",
"Rps26",
"Rps28",
"Rps27a",
"Rps27l",
"Rps4y1",
"Rps6ka4",
"Rptor"
)
# check whether any are missing and remove them
if (length(setdiff(mtor_genes, genes))) {
  print(sprintf("The following mTOR-implicated genes are not present and will be ignored: %s", setdiff(mtor_genes, genes)))
  mtor_genes <- intersect(mtor_genes, genes)
}

g.gemm_bmdm <- rownames(gemm_mono.vs.h_mono)
g.gl261_bmdm <- rownames(gl261_mono.vs.h_mono)
g.both_bmdm <- intersect(g.gemm_bmdm, g.gl261_bmdm)

g.gemm_mg <- rownames(gemm_mg.vs.h_mg)
g.gl261_mg <- rownames(gl261_mg.vs.h_mg)
g.both_mg <- intersect(g.gemm_mg, g.gl261_mg)

g.bmdm.only <- setdiff(g.both_bmdm, g.both_mg)
g.mg.only <- setdiff(g.both_mg, g.both_bmdm)
g.both <- intersect(g.both_mg, g.both_bmdm)
g.all <- union(g.both_mg, g.both_bmdm)

t <- intersect(g.gemm_mg, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GEMM MG model (tumour vs healthy): %i", length(t)
))
if (length(t)) {
  print(gemm_mg.vs.h_mg[t,])
}
print("****************")

t <- intersect(setdiff(g.gemm_mg, g.gemm_bmdm), mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GEMM MG model (tumour vs healthy) and not in GEMM monocytes: %i", length(t)
))
if (length(t)) {
  print(gemm_mg.vs.h_mg[t,])
}
print("****************")

t <- intersect(g.gl261_mg, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GL261 MG model (tumour vs healthy): %i", length(t)
))
if (length(t)) {
  print(gl261_mg.vs.h_mg[t,])
}
print("****************")

t <- intersect(setdiff(g.gl261_mg, g.gl261_bmdm), mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GL261 MG model (tumour vs healthy) and not in GL261 monocytes: %i", length(t)
))
if (length(t)) {
  print(gl261_mg.vs.h_mg[t,])
}
print("****************")


t <- intersect(g.mg.only, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in both MG models (tumour vs healthy) and not in monocytes: %i", length(t)
))
if (length(t)) {
  print(gemm_mg.vs.h_mg[t,])
  print(gl261_mg.vs.h_mg[t,])
}
print("****************")
print("****************")
print("****************")


t <- intersect(g.gemm_bmdm, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GEMM macrophage model (tumour vs healthy): %i", length(t)
))
if (length(t)) {
  print(gemm_mono.vs.h_mono[t,])
}
print("****************")

t <- intersect(setdiff(g.gemm_bmdm, g.gemm_mg), mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GEMM macrophage model (tumour vs healthy) and not in GEMM MG: %i", length(t)
))
if (length(t)) {
  print(gemm_mono.vs.h_mono[t,])
}
print("****************")

t <- intersect(g.gl261_bmdm, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GL261 macrophage model (tumour vs healthy): %i", length(t)
))
if (length(t)) {
  print(gl261_mono.vs.h_mono[t,])
}
print("****************")

t <- intersect(setdiff(g.gl261_bmdm, g.gl261_mg), mtor_genes)
print(sprintf(
  "Number mTOR genes DE in GL261 macrophage model (tumour vs healthy) and not in GL261 MG: %i", length(t)
))
if (length(t)) {
  print(gl261_mono.vs.h_mono[t,])
}
print("****************")

t <- intersect(g.bmdm.only, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in both macrophage models (tumour vs healthy) and not in MG: %i", length(t)
))
if (length(t)) {
  print(gemm_mono.vs.h_mono[t,])
  print(gl261_mono.vs.h_mono[t,])
}
print("****************")

t <- intersect(g.both, mtor_genes)
print(sprintf(
  "Number mTOR genes DE in both monocyte models AND both MG models (tumour vs healthy): %i", length(t)
))
if (length(t)) {
  print(gemm_mg.vs.h_mg[t,])
  print(gl261_mg.vs.h_mg[t,])
  print(gemm_mono.vs.h_mono[t,])
  print(gl261_mono.vs.h_mono[t,])
}
print("****************")

```
 
This demonstrates that the GL261 model has a great deal more DE genes overlapping with the mTOR signature that are present in MG but not BMDM. Most of these are upregulated, including all the Rps genes.

Let's also generate a few boxplots for these genes, to visualise the actual difference in counts. We'll generate separate outputs for the two models.
 
```{r, fig.width=12}
# reformat so that each row is an observation
mtor_meta <- meta[meta$model %in% c('healthy', 'gemm'),]
mtor_data <- as.data.frame(t(data[mtor_genes, rownames(mtor_meta)]))
mtor_data$group <- factor(paste(mtor_meta$cell_type, mtor_meta$disease, sep='_'))
meltData <- melt(mtor_data)
p <- ggplot(meltData, aes(x = group, y = value, fill = group))
p <- p + geom_boxplot() + geom_jitter(mapping = aes(alpha=0.2, stroke=0), width = 0.3, show.legend = F) + scale_fill_manual(values=c("#ff9999", "#ff1a1a", "#80dfff", "#1a1aff"))
p <- p + theme(axis.text.x = element_blank()) + labs(y="Gene counts")
p + facet_wrap(~variable, scale="free") + ggtitle("GEMM model")

# grid.newpage()

mtor_meta <- meta[meta$model %in% c('healthy', 'gl261'),]
mtor_data <- as.data.frame(t(data[mtor_genes, rownames(mtor_meta)]))
mtor_data$group <- factor(paste(mtor_meta$cell_type, mtor_meta$disease, sep='_'))
meltData <- melt(mtor_data)
p <- ggplot(meltData, aes(x = group, y = value, fill = group))
p <- p + geom_boxplot() + geom_jitter(mapping = aes(alpha=0.2, stroke=0), width = 0.3, show.legend = F) + scale_fill_manual(values=c("#ff9999", "#ff1a1a", "#80dfff", "#1a1aff"))
p <- p + theme(axis.text.x = element_blank()) + labs(y="Gene counts")
p + facet_wrap(~variable, scale="free") + ggtitle("GL261 model")

```

This further demonstrates that the effects are more clearly observed in the GL261 model. We can see several genes that we might expect to be tagged as DE in _both_ groups, but which only show up in one (or neither), for example Rps25 in the GEMM model appears to show a strong upregulation in both MG and BMDM. This must be due to the estimation of dispersion - the large number of comparisons being made means that effects need to be very strong to be picked out as significant.

 
Finally, we output the DE gene lists to text files.
 
```{r}
out.dir <- getOutputDir('mtor')
```