---
title: "tcga_survival_metadata"
author: "Gabriel Rosser"
date: "5 January 2018"
output: html_document
---

```{r setup, include=FALSE}
require("knitr")

knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = normalizePath('..'))
```

# Load the data

First we load the TCGA data, complete with the mTOR, MG and BMDM gene signature scores.

```{r}
# This just defines the data directories
source('_settings.R')
in.file <- file.path(data.dir, 'bowman_tcga', 'bowman_data_with_scores.csv')
data <- read.csv(in.file)
```

# Exploratory

Plot survival vs each of the scores.

```{r}
plot(data$mtor_score_z, data$survival)
plot(data$mg_score_z, data$survival)
plot(data$bmdm_score_z, data$survival)
```

# Statistical model

Now we define and fit a very simple statistical model with no interaction terms, just the intercept and two parameters for mTOR and MG signatures.

```{r}
formula <- survival ~ mtor_score_z + mg_score_z
fit <- lm(formula, data)
summary(fit)
```

Only the intercept is significant.

Now we try a more fancy model with an interaction term

```{r}
formula <- survival ~ mtor_score_z * mg_score_z
fit <- lm(formula, data)
summary(fit)
```

Still nothing.
