---
title: "Testing for batch effects in hGIC methylation data"
author: "Gabriel Rosser"
date: "12 February 2019"
output: html_document
---

```{r setup, include=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = normalizePath('..'))

knitr::opts_chunk$set(echo = TRUE)

require(limma)
require("ChAMP")
require("minfi")
require("wateRmelon")
require("data.table")
require(sva)
require(RColorBrewer)
require(openxlsx)
require(ks)
require("ggplot2")
require("reshape")

# this specifies the normalisation algorithm to use
norm.fun <- 'swan'
nhid_to_patient_id <- list("NH16-2214"="052",
"NH16-2806"="061",
"NH16-270"="026",
"NH16-616"="030",
"NH16-1976"="049",
"NH16-1574"="044",
"NH16-677"="031",
"NH15-1877"="018",
"NH15-2101"="019",
"NH16-2255"="054",
"NH16-2063"="050",
"NH15-1661"="017"
)

```

## Load data and prepare metadata

Load the data from the raw `idat` files. We'll load in two blocks: FFPE and cell culture. This ensures that all samples are normalised together for each preparation type.

```{r echo=FALSE}
source('_settings.R')
source("methylation/loader.R")

ffpe_samples <- c(
  'NH15-1877',
  'NH15-2101',
  'NH16-616',
  'NH16-677',
  'NH15-1661',
  'NH16-2063',
  'NH16-2255',
  'NH16-2806',
  'NH16-270',
  'NH16-2214'
)

ffpe_base.dirs <- c(
  file.path(data.dir.raid, 'methylation', '2016-06-10_brandner'),
  file.path(data.dir.raid, 'methylation', '2017-01-17_brandner'),
  file.path(data.dir.raid, 'methylation', '2017-02-09_brandner')
)

cc_samples <- c(
  "GBM018_P12",
  "GBM018_P10",
  "DURA018_NSC_N4_P4",
  "DURA018_NSC_N2_P6",
  "GBM019_P4",
  "GBM019_P3n6",
  "DURA019_NSC_N8C_P2",
  "DURA019_NSC_N5C1_P2",
  "GBM030_P9",
  "GBM030_P5",
  "DURA030_NSC_N16B6_P1",
  "DURA030_NSC_N9_P2",
  "GBM031_P7",
  "GBM031_P4",
  "DURA031_NSC_N44B_P2",
  "DURA031_NSC_N44F_P3",
  "GBM017_P3",
  "GBM017_P4",
  "DURA017_NSC_N3C5_P4",
  "GBM050_P7n8",
  "GBM050_P9",
  "DURA050_NSC_N12_P3",
  "DURA050_NSC_N16_P4",
  "GBM054_P4",
  "GBM054_P6",
  "DURA054_NSC_N3C_P2",
  "DURA054_NSC_N2E_P1",
  "GBM061_P3",
  "GBM061_P5",
  "DURA061_NSC_N4_P2",
  "DURA061_NSC_N1_P3n4",
  "GBM026_P8",
  "GBM026_P3n4",
  "DURA026_NSC_N31D_P5",
  "GBM052_P6n7",
  "GBM052_P4n5",
  "DURA052_NSC_N4_P3",
  "DURA052_NSC_N5_P2"
)

cc_base.dirs <- c(
  file.path(data.dir.raid, 'methylation', "2017-05-12"),
  file.path(data.dir.raid, 'methylation', "2016-12-19_ucl_genomics"),
  file.path(data.dir.raid, 'methylation', "2017-09-19"),
  file.path(data.dir.raid, 'methylation', "2018-01-12"),
  file.path(data.dir.raid, 'methylation', "2017-08-23"),
  file.path(data.dir.raid, 'methylation', "2018-04-09")
)

lst <- get_filenames_from_batches(ffpe_base.dirs, ffpe_samples)

res <- process_idats(lst$in.files, lst$snames, norm.fun=norm.fun)

ffpe_beta <- res$beta
ffpe_m <- MfromBeta(ffpe_beta)

ffpe_meta <- data.frame(row.names = lst$snames)
ffpe_meta$batch <- lst$batches
ffpe_meta$patient_id <- nhid_to_patient_id[rownames(ffpe_meta)]

lst <- get_filenames_from_batches(cc_base.dirs, cc_samples)

cc_res <- process_idats(lst$in.files, lst$snames, norm.fun=norm.fun, force = T)

cc_beta <- cc_res$beta
cc_m <- MfromBeta(cc_beta)

cc_meta <- data.frame(row.names = lst$snames)
cc_meta$batch <- lst$batches
cc_meta$patient_id <- gsub("[^0-9]*([0-9]{3}).*", "\\1", rownames(cc_meta), perl=T)
cc_meta$cell_type <- "iNSC"
cc_meta[grep('GBM', rownames(cc_meta)), 'cell_type'] <- 'GIC'

```
## Check for batch effects

We're interested in whether batch effects are present in either of the blocks of data. We start by looking at the distribution of M values. Do these show any obvious correlation with batch?

```{r batch_effects}
this_meta <- data.frame(ffpe_meta)
eval.points <- seq(from = -10, to = 10, by = 0.05)
fhat <- apply(ffpe_m, MARGIN = 2, FUN = function(x) kde(x, eval.points = eval.points))
df <- data.frame(lapply(fhat, FUN = function(x) x$estimate))
colnames(df) <- this_meta$patient_id
rownames(this_meta) <- this_meta$patient_id

df_long <- melt(cbind(eval.points, df), id = 'eval.points')
df_long$batch <- this_meta[df_long$variable, 'batch']

ggplot(df_long, aes(x=eval.points, y=value, hue=variable)) + geom_line(aes(color=batch), size=1., alpha=0.7) + labs(title="M value distribution in FFPE", x='M value', y='Density')

this_meta <- data.frame(cc_meta)
fhat <- apply(cc_m, MARGIN = 2, FUN = function(x) kde(x, eval.points = eval.points))
df <- data.frame(lapply(fhat, FUN = function(x) x$estimate))
colnames(df) <- this_meta$patient_id

pid_to_batch <- this_meta[this_meta$cell_type == 'GIC',]
pid_to_batch <- pid_to_batch[!duplicated(pid_to_batch$patient_id),]
rownames(pid_to_batch) <- pid_to_batch$patient_id

df_long <- melt(cbind(eval.points, df), id = 'eval.points')
df_long$batch <- pid_to_batch[df_long$variable, 'batch']

ggplot(df_long, aes(x=eval.points, y=value, hue=variable)) + geom_line(aes(color=batch), size=1., alpha=0.7) + labs(title="M value distribution in cell culture", x='M value', y='Density')
```

It appears not, certainly for the cell culture results. FFPE results are trickier to interpret, because the distributions appear more variable. This may be due to differing tumour purity and cell type content?

Let's plot a PCA of the most variable probes.

```{r batch_effects_pca}
n_probe <- 5000
ffpe_mad <- as.data.frame(rowMads(ffpe_m))
rownames(ffpe_mad) <- rownames(ffpe_m)
ix <- rownames(ffpe_mad)[order(ffpe_mad, decreasing = T)[1:n_probe]]
ffpe_pca <- data.frame(prcomp(ffpe_m[ix,])$rotation)
ffpe_pca$batch <- ffpe_meta[rownames(ffpe_pca), 'batch']
ggplot(ffpe_pca, aes(x=PC1, y=PC2, hue=batch)) + geom_point(aes(color=batch)) + labs(title=sprintf("PCA of top %d FFPE M values", n_probe))

this_cc <- cc_m[, cc_meta$cell_type == 'GIC']
cc_mad <- as.data.frame(rowMads(this_cc))
rownames(cc_mad) <- rownames(this_cc)
ix <- rownames(cc_mad)[order(cc_mad, decreasing = T)[1:n_probe]]
cc_pca <- data.frame(prcomp(this_cc[ix,])$rotation)
cc_pca$batch <- cc_meta[rownames(cc_pca), 'batch']
ggplot(cc_pca, aes(x=PC1, y=PC2, hue=batch)) + geom_point(aes(color=batch)) + labs(title=sprintf("PCA of top %d cell culture M values", n_probe))

```


